Efficient pedigree recording for fast population genetics simulation

### 2017-11-15 forwards_paper.tex
changed title; begun simplify alg writeup

\title{Efficient forwards-time simulation with tree sequences}
\author{Jerome Kelleher,
        Kevin R. Thornton,
        Jaime Ashander, and
        Peter L. Ralph}
\maketitle

@@

Modern computing power easily allows purely \emph{demographic} simulations
of a population having even hundreds of millions of individuals,
even though theory tells us that a population of size $N$
must be run for many multiples of $N$ generations to produce stable genetic patterns.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must somehow produce at the end data for each individual on a genomic scale.
As samples of most species genomes harbor tens or hundreds of millions of variant sites,
naively carrying full genotypes for even modest numbers of individuals through a simulation 
becomes quickly prohibitive.

@@

A \emph{tree sequence} is an encoding for the sequence of correlated trees 
that describes the history of a sexual population.
It is efficient because branches that are shared by adjacent trees are stored once, 
rather than repeatedly for each tree.
The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
while information about variants are recorded as \emph{sites} and \emph{mutations}.
The format for these four tables is given in more detail in the Methods;
here we explain conceptually,
using the example of Figure~\ref{fig:example_tree_sequence}.

@@

A minimal genealogy such as in Figure~\ref{fig:wf-trees}
has fewer edges on which to place mutations.
There are many more advantages to simplifying the genealogies to this point.
Computing this minimal representation of the
edges output by a forward-time simulation is an instance of a more
general problem that cover in the next section.

@@

Here we describe the simplification algorithm.
\plr{I am trying it out to see what it looks like.}

\begin{taocpalg}{S}{Simplify a tree sequence}
{Simplifies the input tree sequence.
}

\algstep{S0.}{Initialize.}{
    Set the samples.
}

@@

\algstep{S2.}{Process parent edges.}{Loop over edges associated with this parent, removing ancestry from each child.
}

@@

This uses
\begin{taocpalg}{R}{Remove ancestry}
    {Remove ancestry on a segment from an ancestor.}

\end{taocpalg}
and
\begin{taocpalg}{M}{Merge ancestors}
    {Merge a collection of segments and create a new ancestor.}

>>>>

\title{Efficient pedigree recording for fast population genetics simulation}
\author{Jerome Kelleher,
        Kevin R. Thornton,
        Jaime Ashander, and
        Peter L. Ralph}
\maketitle

@@

Modern computing power easily allows simulations of birth, death and reproduction
in a population having even hundreds of millions of individuals,
even though theory tells us that a population of size $N$
must be run for many multiples of $N$ generations to produce stable genetic patterns.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must somehow produce at the end data for each individual on a genomic scale.
As samples of most species genomes harbor tens or hundreds of millions of variant sites,
naively carrying full genotypes for even modest numbers of individuals through a simulation 
becomes quickly prohibitive.

@@

A \emph{tree sequence} is an encoding for the sequence of correlated trees,
such as those that describe the history of a sexual population.
It is efficient because branches that are shared by adjacent trees are stored once, 
rather than repeatedly for each tree.
The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
while information about variants are recorded as \emph{sites} and \emph{mutations}.
The format for these four tables is given in more detail in the Methods;
here we explain conceptually,
using the example of Figure~\ref{fig:example_tree_sequence}.

@@

A minimal genealogy such as in Figure~\ref{fig:wf-trees}
has fewer edges on which to place mutations.
There are many more advantages to simplifying the genealogies to this point.
Computing this minimal representation of the
edges output by a forward-time simulation is an instance of a more
general problem that we cover in the next section.

@@

Here we describe the simplification algorithm.
The flow is close to our implementation,
but the description below uses set operations,
while in the implementation,
all ancestors are maintained as linked lists of segments
ordered by left endpoint.

\begin{taocpalg}{S}{Simplify a tree sequence}
{Simplifies the input tree sequence:
    input consists of 
    a list of sample IDs, $S$, of length $n$,
    a list of edges, $\mathcal{E_I}$,
    and a list of nodes $\mathcal{N_I}$.
    The genome length is $L$.
    The output will be stored in a list of edges $\mathcal{E_O}$ 
    and nodes $\mathcal{N}_O$.
    This also maintains the list $N$ that gives the mapping 
    from $\mathcal{N_I}$ to $\mathcal{N_O}$,
    so that if $M[p] = u$ then the $p^\text{th}$ ancestor in $N_I$
    is the same as the $u^\text{th}$ ancestor in $N_0$.
    The internal state is $\mathcal{A}$, the list of ancestors
    (which are collections of ancestral segments).
}

\algstep{S0.}{Initialize.}{
    Set $\mathcal{A}[j] \leftarrow (0, L, j)$ 
    and $\mathcal{N}_O[j] \leftarrow \mathcal{N}_I[S[j]]$ for $0 \le j < N$.
    Set $Q \leftarrow \emptyset$,
    and $M[p] \leftarrow -1$ for $0 \le p < \text{length}(N_I)$.
}

@@

\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} 
    on each edge corresponding to the current parent.
}

@@

The first step is, for each edge $(\ell, r, p, c)$,
simply to replace any ancestry segment $[a, b)$ held by ancestor $c$
with $[a,b) \setdiff [\ell, r)$ 
(deleting the segment or splitting in two if necessary),
and adding the removed segment $[a,b) \cap [\ell, r)$ to the merge queue $Q$:
\begin{taocpalg}{R}{Remove ancestry}
    {Given an edge $(\ell, r, p, c)$,
    and a list $\mathcal{A}[c]$ of the $m$ ancestral segments corresponding to $c$,
    remove segments overlapping $[\ell, r)$ from $\mathcal{A}[c]$,
    and add those overlapping segments to $Q$.
    }

    \algstep{R1.}{Segment loop.}{Loop over ancestral segments in $\mathcal{A}[c]$:
    if $j=m$, terminate; else let $(a, b, u) = \mathcal{A}[c][j]$.
    }

    \algstep{R2.}{Add to merge queue.}{If $[a,b) \cap [\ell,r) \neq \emptyset$,
    append $(\max(a,\ell), \min(b,r), u)$ to $Q$.
    }

    \algstep{R3.}{Remove ancestry.}{If $[a,b) \cap [\ell,r) != \emptyset$,
    delete $\mathcal{A}[c][j]$ and insert in its place
    $(\min(a,\ell), \max(a,\ell), u)$ (if $a<\ell$) 
    and/or $(\min(b,r), \max(b,r), u)$ (if $r<b$).
    }

Next we must merge these ancestry segments,
creating the new ancestor $p$,
outputting edges and a node if coalescence has occurred.
\begin{taocpalg}{M}{Merge ancestors}
    {Merge a collection of segments and create a new ancestor.
    This operates on the merge queue $Q$ above,
    which is a list of ancestral segments,
    and creates the new ancestor $\mathcal{A}[p]$.}

    \algstep{M1.}{Segment loop.}{While $\mathrm{length}(Q)>0$:}

    \algstep{M2.}{Find next segment overlap boundaries}{
    Let $\ell \leftarrow \min\{x.\ell : x \in Q\}$
    and $H \leftarrow \{x \in Q : x.\ell = \ell\}$.
    Also let $\ell_+ \leftarrow \min\{x.\ell : x \in Q \setdiff H\}$
    and $r = \min(\ell_+, \min\{x.r : x \in H\})$.
    }

    \algstep{M3.}{Copy single segment.}{If $\mathrm{length}(H) == 1$,
    set $\alpha \leftarrow (\ell, r, H[0].n)$.
    If also $p \in S$, then append $(\alpha.\ell, \alpha.r, N[p], \alpha.c)$ to $\mathcal{E}_O$
    and then set $\alpha.c \leftarrow M[p]$.
    Go to \algref{M7}.
    }

    \algstep{M4.}{Coalescence head.}{If $(h \leftarrow \mathrm{length}(H)) > 1$,
    if $M[p] == -1$ set $M[p] \leftarrow \text{length}(N_O)$
    and append $N_I[p]$ to $N_O$.
    Set $\alpha \leftarrow (\ell, r, M[p])$.
    }

    \algstep{M6.}{Record edges.}{
    For each $x \in H$, append $(\ell, r, M[p], x.n)$ to $\mathcal{E}_O$.
    }

    \algstep{M7.}{Update $Q$.}{
    Set $Q \leftarrow Q \setdiff \{x \in Q : x.r == r\}$
    and then set $x.\ell = \ell$ for all remaining $x \in Q$.
    }

    \algstep{M8.}{Add segment to ancestor.}{
    Append $\alpha$ to $\mathcal{A}[p]$ and return to \algref{M1}.
    }
